`You Draw It': Implementation of visually fitted trends with \texttt{r2d3}

### 2022-07-05 2022-sdss-you-draw-it-manuscript.Rmd
Add references and example of mental operations

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = F, 
                      eval = T, 
                      fig.path="./images/", 
                      fig.env="figure*", 
                      fig.align = "center")
library(readr)
library(tidyverse)
library(knitr)
library(bibtex)
```

@@

We all use statistical graphics, but how do we know that the graphics we use are communicating properly?
Through experimentation, graphical testing methods allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.
All of these types of tests require different levels of use and manipulation of the information being presented in the chart.
Efforts in the field of statistical graphics have developed graphical testing tools and methods, such as the lineup protocol \cite{buja2009statistical}. 
The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.

\subsection{Measuring Patterns and Trends}

One such aspect of interest is the ability to identify and detect trends in data. 
Our visual system is naturally built to look for structure and identify patterns. 
For instance, points going down from left to right indicates a negative correlation between the $x$ and $y$ variables.

Initial studies in the 20th century explored the use of fitting lines by eye through a set of points \cite{finney1951subjective,mosteller1981eye}. Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. Recently, Ciccione and Dehaene \shortcite{ciccione2021can} conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.

In 2015, the New York Times introduced an interactive feature, called 'You Draw It' \cite{aisch2015you}, where readers input their own assumptions about various metrics and compare how these assumptions relate to reality. The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their mouse.

What remains to be determined is how we can compare our intuitive visual sense of patterns to those determined by statistical methods.

\subsection{Research Objectives}
The goal of this research is to implement ‘You Draw It’, adapted from the New York Times feature, as a way to measure the patterns we see in data. Here, we provide technical details of the software development, utilizing interactive graphics in R. We then share results from our study which validates 'You Draw It' as a method for graphical testing and apply an appropriate data analysis method to the participant data.

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = F, 
                      eval = T, 
                      fig.path="./images/", 
                      fig.env="figure*", 
                      fig.align = "center")
library(readr)
library(tidyverse)
library(knitr)
library(bibtex)
theme_set(theme_bw())

library(readxl)
library(lubridate)
```

@@

We all use statistical graphics, but how do we know that the graphics we use are communicating \svp{effectively}?
Through experimentation, graphical testing methods allow researchers to conduct studies \svp{designed to understand how we perceive graphics and perform graphical tasks} such as differentiation, prediction, estimation, and extrapolation. 
\svp{Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.}
<!-- All of these types of tests require different levels of use and manipulation of the information being presented in the chart. -->
<!-- Efforts in the field of statistical graphics have developed graphical testing tools and methods, such as the lineup protocol \cite{buja2009statistical}.  -->
<!-- The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics. -->
\svp{In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics. 
We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.}

\subsection{Measuring Patterns and Trends}
\svp{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
These charts rely on our }ability to identify and detect trends in data. 
Our visual system is naturally built to look for structure and identify patterns, \svp{including patterns and trends over time}. 
\svp{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time will likely perform several cognitive operations. First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points. Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability. Finally, the viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}

```{r gas-prices, echo = F, message = F, warning = F}
#| fig-cap: "Weekly average gas prices in the United States, 2019-June 2022. Additional mental operations a viewer might perform while looking at the plot are annotated in grey. \\cite{usenergyinformationadministrationWeeklyAllGrades2022}."
#| out-width: "\\columnwidth"
#| label: gas-prices
#| fig-res: 300
#| fig-width: 8
#| fig-height: 4
#| fig-env: figure

gas <- read_xls("EMM_EPM0_PTE_NUS_DPGw.xls", skip = 3, sheet = 2) %>%
  set_names(c("Date", "Price"))

annotations <- tibble(
  xmin = as.POSIXct(ymd(c("2020-03-15", "2022-02-26"))),
  xmax = as.POSIXct(ymd(c("2020-05-30", "2022-06-29"))),
  ymin = -Inf,
  ymax = Inf,
  label = c("COVID lockdowns", "War in Ukraine")
)
ggplot(gas, aes(x = Date, y = Price)) + 
  geom_rect(data = annotations, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), alpha= 0.2, fill = "grey", inherit.aes = F) + 
  geom_text(data = annotations, aes(x = xmin, y = 1, label = label), color = "grey") + 
  ylab("Price per Gallon ($)") + 
  ggtitle("Weekly Average Gas Prices in the United States, 2019 - 2022", subtitle = "US Energy Information Administration") + 
  geom_point() + 
  scale_x_datetime(limits = as.POSIXct(ymd("2019-01-01", "2022-06-29"))) + 
  geom_line(se = F, color = "grey", method = "loess", span = .1, alpha = .5, stat = "smooth", size = 1.5)
```

Initial studies in the 20th century explored the use of fitting lines by eye through a set of points \cite{finney1951subjective,mosteller1981eye}. Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit was suitable, then drawing the line through the set of points. Recently, Ciccione and Dehaene \shortcite{ciccione2021can} conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.\svp{XXX and found what? Describe the results XXX}

\svp{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.} In 2015, the New York Times introduced an interactive feature, called 'You Draw It' \cite{aisch2015you}, where readers input their own assumptions about various \svp{metrics of news interest} and compare these assumptions to reality. The New York Times team used Data Driven Documents (D3), \svp{a JavaScript library} that allows readers to \svp{interact with a chart directly by} drawing a line on their computer screen with a mouse. \svp{Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.}

\svp{We set out to implement} ‘You Draw It’, adapted from the New York Times feature, as a way to measure the patterns we see in data. 
Here, we provide technical details of the software development, utilizing interactive graphics in R\cite{r-software}. 
We then share results from our study which validates 'You Draw It' as a method for graphical testing \svp{on linear trends} and apply an appropriate data analysis method to the participant data. \svp{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}

### 2022-07-05 2022-sdss-you-draw-it-manuscript.tex
Add references and example of mental operations

\begin{document}
\maketitle
\begin{abstract}
How do statistical regression results compare to intuitive, visually
fitted results? Fitting lines by eye through a set of points has been
explored since the 20th century. Common methods of fitting trends by eye
involve maneuvering a string, black thread, or ruler until the fit is
suitable, then drawing the line through the set of points. In 2015, the
New York Times introduced an interactive feature, called `You Draw It,'
where readers are asked to input their own assumptions about various
metrics and compare how these assumptions relate to reality. This
research is intended to implement `You Draw It,' adapted from the New
York Times, as a way to measure the patterns we see in data.
\end{abstract}

@@

We all use statistical graphics, but how do we know that the graphics we
use are communicating properly? Through experimentation, graphical
testing methods allow researchers to conduct studies geared at
understanding human ability to conduct tasks related to the perception
of statistical charts such as differentiation, prediction, estimation,
and extrapolation. All of these types of tests require different levels
of use and manipulation of the information being presented in the chart.
Efforts in the field of statistical graphics have developed graphical
testing tools and methods, such as the lineup protocol
\cite{buja2009statistical}. The advancement of graphing software
provides the tools necessary to develop new methods of testing
statistical graphics.

@@

One such aspect of interest is the ability to identify and detect trends
in data. Our visual system is naturally built to look for structure and
identify patterns. For instance, points going down from left to right
indicates a negative correlation between the \(x\) and \(y\) variables.

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points
\cite{finney1951subjective,mosteller1981eye}. Common methods of fitting
trends by eye involved maneuvering a string, black thread, or ruler
until the fit is suitable, then drawing the line through the set of
points. Recently, Ciccione and Dehaene \shortcite{ciccione2021can}
conducted a comprehensive set of studies investigating human ability to
detect trends in graphical representations from a psychophysical
approach.

In 2015, the New York Times introduced an interactive feature, called
`You Draw It' \cite{aisch2015you}, where readers input their own
assumptions about various metrics and compare how these assumptions
relate to reality. The New York Times team utilizes Data Driven
Documents (D3) that allows readers to predict these metrics through the
use of drawing a line on their computer screen with their mouse.

What remains to be determined is how we can compare our intuitive visual
sense of patterns to those determined by statistical methods.

\subsection{Research Objectives}

The goal of this research is to implement `You Draw It,' adapted from
the New York Times feature, as a way to measure the patterns we see in
data. Here, we provide technical details of the software development,
utilizing interactive graphics in R. We then share results from our
study which validates `You Draw It' as a method for graphical testing
and apply an appropriate data analysis method to the participant data.

@@

When completing the graphical task, users are shown an interactive
scatter-plot (Figure 1) along with the prompt, ``Use your mouse to fill
in the trend in the yellow box region.'' The yellow box region moves
along as the user draws their trend-line, providing a visual cue which
indicates where the user still needs to complete a trend line. After the
entire domain has been visually estimated or predicted, the yellow
region disappears, indicating the participant has completed the task.
Prior to study participation, example gifs are shown and users are asked
to complete practice plots to ease the learning curve of the
task.\footnote{Visit \url{emily-robinson.shinyapps.io/can-you-draw-it}
  for a test applet.}

@@

Parameters for aesthetic design choices are defined in a list of options
and \texttt{r2d3} passes these to the D3.js code. For instance, we can
specify the buffer space allowed for the \(x\) and \(y\) axes to avoid
users to anchor their lines to the axes limits.\footnote{For
  \texttt{D3.js} source code, visit
  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}

>>>>

\begin{document}
\maketitle
\begin{abstract}
How do statistical regression results compare to intuitive, visually
fitted results? Fitting lines by eye through a set of points has been
explored since the 20th century. Common methods of fitting trends by eye
involve maneuvering a string, black thread, or ruler until the fit is
suitable, then drawing the line through the set of points. In 2015, the
New York Times introduced an interactive feature, called `You Draw It,'
where readers are asked to input their own assumptions about various
metrics and compare how these assumptions relate to reality. This
research is intended to implement `You Draw It', adapted from the New
York Times, as a way to measure the patterns we see in data.
\end{abstract}

@@

We all use statistical graphics, but how do we know that the graphics we
use are communicating {\textcolor{RedOrange}{effectively}}? Through
experimentation, graphical testing methods allow researchers to conduct
studies
{\textcolor{RedOrange}{designed to understand how we perceive graphics and perform graphical tasks}}
such as differentiation, prediction, estimation, and extrapolation.
{\textcolor{RedOrange}{Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.}}
{\textcolor{RedOrange}{In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics. 
We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.}}

@@

{\textcolor{RedOrange}{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
These charts rely on our }}ability to identify and detect trends in
data. Our visual system is naturally built to look for structure and
identify patterns,
{\textcolor{RedOrange}{including patterns and trends over time}}.
{\textcolor{RedOrange}{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time will likely perform several cognitive operations. First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points. Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability. Finally, the viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}}

\begin{figure}

{\centering \includegraphics[width=\columnwidth]{./images/gas-prices-1}

}

\caption{Weekly average gas prices in the United States, 2019-June 2022. Additional mental operations a viewer might perform while looking at the plot are annotated in grey. \cite{usenergyinformationadministrationWeeklyAllGrades2022}.}\label{fig:gas-prices}
\end{figure}

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points
\cite{finney1951subjective,mosteller1981eye}. Common methods of fitting
trends by eye involved maneuvering a string, black thread, or ruler
until the fit was suitable, then drawing the line through the set of
points. Recently, Ciccione and Dehaene \shortcite{ciccione2021can}
conducted a comprehensive set of studies investigating human ability to
detect trends in graphical representations from a psychophysical
approach.{\textcolor{RedOrange}{XXX and found what? Describe the results XXX}}

{\textcolor{RedOrange}{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.}}
In 2015, the New York Times introduced an interactive feature, called
`You Draw It' \cite{aisch2015you}, where readers input their own
assumptions about various
{\textcolor{RedOrange}{metrics of news interest}} and compare these
assumptions to reality. The New York Times team used Data Driven
Documents (D3), {\textcolor{RedOrange}{a JavaScript library}} that
allows readers to
{\textcolor{RedOrange}{interact with a chart directly by}} drawing a
line on their computer screen with a mouse.
{\textcolor{RedOrange}{Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.}}

{\textcolor{RedOrange}{We set out to implement}} `You Draw It', adapted
from the New York Times feature, as a way to measure the patterns we see
in data. Here, we provide technical details of the software development,
utilizing interactive graphics in R\cite{r-software}. We then share
results from our study which validates `You Draw It' as a method for
graphical testing {\textcolor{RedOrange}{on linear trends}} and apply an
appropriate data analysis method to the participant data.
{\textcolor{RedOrange}{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}}

@@

When completing the graphical task, users are shown an interactive
scatter-plot (Figure 1) along with the prompt, ``Use your mouse to fill
in the trend in the yellow box region.'' The yellow box region moves
along as the user draws their trend-line, providing a visual cue which
indicates where the user still needs to complete a trend line. After the
entire domain has been visually estimated or predicted, the yellow
region disappears, indicating the participant has completed the task.
Prior to study participation, example gifs are shown and users are asked
to complete practice plots to ease the learning curve of the task.
\footnote{Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a
  test applet.}

@@

Parameters for aesthetic design choices are defined in a list of options
and \texttt{r2d3} passes these to the D3.js code. For instance, we can
specify the buffer space allowed for the \(x\) and \(y\) axes to avoid
users to anchor their lines to the axes limits. \footnote{For
  \texttt{D3.js} source code, visit
  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}
